Leadership across cultures and contexts: an exploratory analysis

### 2020-04-29 leadership_across_cultures_contexts.Rmd
Tweaked resuls text on clusters

Across human societies, the influence and agendas of leaders stands to substantially shape group structure and group outcomes. The structure of groups, their goals, and the demands of leaders can favor particular leadership qualities as well as the costs and benefits to individuals involved. Cultural norms, capacities of competing leaders, inter-group dynamics, and follower preferences can also impact leader agendas. The interactions between leader strategies, pressures of followership, and group dynamics are likely to strongly contribute to organizational variation across human populations including social and economic trajectories [@wiessner_risk_1982] and cultural and biological evolutionary processes [@henrich_big_2015].

<!-- NOT SURE WHERE THIS SENTANCE SHOULD GO OR COULD BE CUT -->
Within every population there are particular contexts in which one or a few individuals maintain a disproportionate level of influence in the group, function to facilitate group goals, and provides services to the group.

Much scientific work across psychology, sociology, political science, and managerial studies has focused on trying to uncover the interrelationships of these dimensions of leadership. Until recently, this large body of work was relatively disconnected from approaches in anthropology and biology. Under a unifying evolutionary framework, social and biological scientists have now produced a sizable body of theory on the evolution and cross-cultural patterning of leadership and followership [e.g., @price_evolution_2014; @henrich_big_2015; @richerson_tribal_2009; @van_vugt_evolutionary_2006; @pietraszewski_evolution_2019; @hooper_theory_2010; @kaplan_evolutionary_2009; @von_rueden_roots_2014; @hagen_leadership_2019], accompanied by a growing body of empirical results [e.g., @bowser_womens_2010; @alami_mothers_2020; @cheng_two_2013; @garfield_investigating_2019; @garfield_evolutionary_2019; @von_rueden_leadership_2014; @von_rueden_sex_2018; @smith_leadership_2016; @smith_obstacles_2018; @lonati_what_2019]. This integration as been informative [see @van_vugt_genes_2020].

There is now well established consensus within behavioral and evolutionary human sciences on the importance of data from the full range of human cultural diversity and systemic biases in much of social science are well known [@henrich_weirdest_2010, *The weirdest people in the world?*]. @henrich_weirdest_2010 in their initial critique of the over-realiance on WEIRD samples emphasized a holistic approach to cultural diversity within and across "telescoping" levels of human societies, but also identified a contrast that, "compares people from modern industrialized societies with those from small-scale societies" (p.62), which has become the most salient interpretation of the paper and call-to-action. Since the seminal WEIRD people paper many social scientists have indeed taken up the challenge to diversify their samples and have sought out "non-WEIRD" participants and databases. However, categorizing populations as WEIRD or non-WEIRD and attempting to charachterize "non-WEIRD" social dynamics to learn about human psychology is a reductionist over-correction and arguably only a marginal scientific improvement from exclusively studying WEIRD populations.^[A Google Scholar search for "non-WEIRD" produced over 1000 results.] Extending the study of leadership to include evolutionary theory and data from diverse, small-scale and traditional societies (i.e., non-WEIRD) can facilitate more generalizable theories and offer applications to shape organizational structure across populations (i.e., WEIRD settings) in ways which are more consistent and less mismatched with human evolutionary history and evolved leader-follower psychologies [@von_rueden_leadership_2015; @garfield_review_2019; @van_vugt_selected_2010]. But such approaches are also at risk of failing to incorporate the message of the larger WEIRD people problem: diversity of cultural, social, and ecological contexts can be strong divers of human behavior and group dynamics.

Much of leadership studies, including evolutionary approaches, have focused on community or political leaders, often overlooking subtilties in the nature of leadership and followership across social domains and levels of social organization, such as among kin groups, economic groups, and sub-groupings of individuals within residential communities. Lastly, while the benefits leaders accrue for their services have received substantial attention -- fitness benefits in particular -- the costs of leadership for both leaders and followers have been relatively under-investigated. Outstanding questions for evolutionary leadership scholars include the degree to which the nature of leadership is associated with population-level variation and in different social contexts and identifying the associated benefits *and* costs both leaders and followers incur.
<!-- CAN PULL SOME COST CITATIONS FROM NHB PAPER -->

<!-- THIS PARAGRAPH CAN BE INCORPORATED ABOVE -->
Evolutionary scholars typically define leaders as individuals who maintain disproportionate influence over decision-making within groups [@von_rueden_leadership_2015], whatever the group may be. Leadership roles are also often associated with social sanctioned rights and responsibilities. Not surprisingly, the functions leaders serve then are often closely linked to group contexts and the salient challenges groups face. Certain types of groups, such as kin groups, economic groups, or political groups, may tend to favor leaders with specific qualities, however there can be substantial variation in the traits of leaders within these groups as well as between them. Adopting the broad definition of leadership given above, parents can also be defined as leaders of nuclear families. Strong community leadership, however, is extremely common cross-culturally including among hunter-gatherers and relatively small-scale societies [@garfield_review_2019; @von_rueden_leadership_2015].

We first briefly review theoretical, ethnographic, and empirical perspectives on the qualities of leaders, the functions they serve, and the costs and benefits of leadership across human societies, including WEIRD and non-WEIRD populations. We then outline the aims and methods of the current cross-cultural study before presenting results and interpretations. Through novel exploratory analyses encompassing the full range of cultural diversity and diverse social contexts we aim to advance leadership studies across disciplines and provide researchers with a foundation for more detailed approaches to human leadership across cultures and contexts.

@@

## The functions of leadership across cultures and contexts
<!-- INTRO -->
Despite the universality of the phenomenon of leadership, the function leaders serve across populations and social contexts is highly diverse and are expected to be closely tied to group structure and group goals [@van_vugt_cognitive_2007]. Among many of the ethnographically described small-scale societies, which generally lack multi-level, institutionalized social and political structures, leadership at the community level is typically informal and ephemeral and leaders emerge contextually when demands of coordination or conflict resolution require a managerial presence [@von_rueden_leadership_2015].

<!-- FUNCTIONS in SSS = CORDINATION AND CONFCLIT RESOLUTION -->
In the absence of institutionalized organizational structure leaders often take on essential roles in group coordination and collective action, including facilitating conflict resolution within or between autonomous households or kin groups, managing individual investments to group-wide projects, and administering sanctions to individuals who do not appropriately contribute to collective actions, comply with group policies, or seriously violate social norms. @glowacki_leadership_2015 draw on empirical, ethnographic, and theoretical models of leader emergence and suggest in small-scale societies institutionalized leadership roles are likely to initially emerge to resolve conflicts between families. @roscoe_social_2009 draws on the ethnography of highland New Guinea in describing the modular structure of human societies and suggests the functions of community leaders in small-scale societies are structured by the joint demands of signaling military formidability to rival groups and facilitating within-group conflict resolution and cooperation. For "big-men" in these societies, in addition to military prowess, "talents for oratory, mediation, and organization were instrumental in reducing their internal conflict and enhancing the ability of their members to coordinate their actions as a group" [@roscoe_social_2009, p. 104].

<!-- GETS INTO QUALITIES BELOW -->
Given demands of community organization and conflict resolution, leadership tends to be context-specific and leaders in such contexts often draw on persuasive capacities and leverage community-based respect, specialized skills or knowledge, and social support to achieve or maintain community influence [@kantner_identifying_2010; @lewis_leaders_1974; @boehm_egalitarian_1993].

<!-- REPETITIVE COMPARED TO ABOVE BUT INCLUDES ADDITIONAL DETAILS about FUNCTIONS AND  MORE CITATIONS, THEORY -->
The theoretical and ethnographic literature suggests these demands most commonly arise in the contexts of collective actions and in conflict resolution [e.g., @glowacki_leadership_2015; @gavrilets_convergence_2016; @hooper_theory_2010; @von_rueden_solving_2015; @henrich_big_2015; @pietraszewski_evolution_2019]. Demands of mediation in within-group conflict resolution, although variable in scale, are probably a universal feature of human groups [@king_origins_2009; @van_vugt_leadership_2008]. Demands of between-group interactions, including both conflict and cooperation are also very common across small-scale societies, including hunter-gatherers [@glowacki_evolutionary_2017; @hames_pacifying_2019; @richerson_cultural_2016; @henrich_evolution_2009; @apicella_social_2012]. Across relatively small-scale societies representing the group and managing both internal and external relationships and conflicts are expected to be the most salient functions of community leaders. Within communities, however, leaders may emerge whenever individuals congregate. The role of leaders in ritual contexts, such as in initiations [@singh_cultural_2017], and in providing social services, including facilitating marriages [@walker_evolutionary_2011] and organizing social activities [@wiessner_food_1996], are also expected to be common functions carried out by a leader.

<!-- MANAGEMENT FUNCTION IN MORE CULTUARLL COMPLEX -->
Increased sedentarization and defensibility of resources is often associated with both social stratification and broader institutionalized managerial roles including stronger authority of community leaders and leadership beyond the community level [@johnson_evolution_1987; @johnson_organizational_1982; @kaplan_evolutionary_2009; @ames_hierarchies_1985; @panchanathan_indirect_2004]. Among horticulturalists and agriculturalists managerial leadership can promote more efficient use of shared resources such as water reserves and cooperative labor [e.g., @von_rueden_leadership_2014]. Pastoralists and coastal populations often face similar pressures concerning grazing lands and fishing access [e.g., @widmer_evolution_1988; @stevens_sherpa_1990]. In the wake of economic intensification, the functions of leadership becomes increasingly associated with managing economic systems and distribution of resources. Defensibility of critical resources creates increased opportunities for resource management and is often associated with increased territoriality and increased pressures of inter-group conflicts [@glowacki_evolutionary_2017]. Community leaders among populations more reliant on domesticated plant foods, livestock, and specific territories are often required to manage military forces and develop both offensive and defensive strategies in warfare, more so than mobile populations that rely on dispersed, less defensible resources [@lopez_evolutionary_2017].

<!-- WEIRD ORGNIZATIONAL STRUCTURE TELLS FUNCTIONS -->
In WEIRD societies, community leaders are primarily heads of governmental and corporate bodies [@bass_handbook_1990]. Other public service, nonprofit leaders and managers include law enforcement and police officers, college presidents and school principles, and grassroots community leaders, to name a few [@martiskainen_role_2017; @azzam_community_2003; @grace_school_2005]. These positions typically are associated with institutionalized responsibilities and specific functions. The structure of such organizations are generally hierarchical with nested bodies and an internal 'chain of command.' At higher levels of large-scale organizations, leaders primarily serve as representatives in inter-group interactions, facilitate the creation and adoption of policies, and serve as figure heads embodying the strategic vision of the organization [@keohane_thinking_2010]. They are not typically directly involved in resolving interpersonal conflicts between individuals, however, they are often the final end-point in a decision making hierarchy and the results of a single decision of these leaders can trickle down through industry or government with global ramifications.

<!-- THEORY ON WEIRD FUNCTIONS -->
Scholars in managerial studies have long discussed the behaviors of leaders and managers have developed multiple models on the functions of leadership. Classic management theorists [e.g., @davis_learning_1951] emphasized the planning, organizing, and controlling functions of leaders in formal organizational and business settings [see @bass_handbook_1990]. These perspectives focused on the rational, mechanistic aspects of group functioning and often overlooked interpersonal dynamics. Later theorists incorporated cognitive and socioemotional components of leadership functions. For example, @selznick_leadership_1957 describes the functions of organization leaders is to define the goals of the organization, create structures for accomplishing group goals, defend institutional integrity, and maintain internal consistency among group members.

<!-- DATA ON WEIRD FUNCTIONS -->
@wofford_factor_1970 provides an early empirical view of managerial functions. A factor analysis of managerial behaviors from over 130 US employees' survey responses identified five interpretable, orthogonal factors: (1) an *order and group achievement* factor, identifying participative, democratic, and organizational planning function; (2) a *personal enhancement oriented* factor, associated with authority and control, (3) a *personal interaction* factor, associated with interpersonal relationships and employee support, (4) a *security and maintenance* factor, associated with self-preservation, and (5) a *dynamic and achievement oriented* factor, associated with aggressive, self-assured actions. Wofford's analysis provided a foundation for future empirical studies of leadership and management functions. He contrasted factor one type management, described as the "professional administrator" who effectively collaborates and integrates team members, from factors two and five, described as "autocratic leaders" strongly motivated by personal power and a willingness to rely on aggressive tactics [@wofford_factor_1970].

<!-- COMPARATIVE DATA ON WEIRD FUNCTIONS -->
More recently, many managerial scholars have adopted integrative and comparative approaches to better understand cultural diversity in leadership systems. @behrendt_integrative_2017 draw on empirical results from the managerial literature and suggest the behaviors of leaders across managerial contexts are classifiable as either *task-oriented*, including enhancing understanding, strengthening motivation and facilitating implementation, or *relation-oriented*, including fostering coordination, promoting cooperation and mobilizing resources. This framework is intended to provide a more fine-grained and externally valid view of leadership functions, compared to traditional models of transactional leadership, which tend to oversimplify dimensions of leader-follower dynamics. More integrated and holistic models have also allowed managerial scholars to develop comparative theoretical frameworks for leader functioning. Culturally Endorsed Implicit Leadership Theory (CLT), suggests cultural variation in social norms has the potential to impact follower preferences for leadership styles [@moore_literature_2018]. The Global Leadership and Organizational Behavior Effectiveness (GLOBE) project is a large-scale international research program aimed at understanding leader effectiveness and cultural variation across large organizations. The GLOBE project has produced multiple cross-national data sets on leadership and hundreds of publications. For example, *The GLOBE Study of CEO Leadership Behavior and Effectiveness in 24 Countries* draws on data from over 1,000 CEOs and over 5,000 senior executives from corporations in varied industries across 24 countries [@house2013strategic]. The goals of the GLOBE CEO Study were to address (1) how national culture influences expectations for the behaviors of leaders, (2) which behaviors of CEOs are typically associated with organizational success, (3) differences between high-functioning and low-functioning CEOs, and (4) the importance of CEO behaviors aligning with societal expectations. Through path analysis the study concluded that cultural values indirectly predict CEO leadership behavior through culturally endorsed leadership expectations. That is, leaders moderate their behavior through culturally learned values and leadership-specific values.

The GLOBE CEO Study defines six "global leadership behaviors" developed from 21 "primary leadership behaviors." Among the global leadership behaviors, the GLOBE CEO Study found that charismatic leadership behavior is consistently the most important behavioral trait regarding firm performance and dedication of top management team (TMT) members, followed by team-oriented behavior and humane-oriented behavior. Participative leadership was moderately related to TMT dedication, however was not related to firm performance. The most successful CEOs are those who are rated as more visionary, performance oriented, and decisive. Furthermore, results strongly suggest the fit between a CEOs behavioral profile and the expectations of leadership within a society strongly predict TMT dedication and firm performance. By higher-order features of leadership behavior and situating leadership functions within more detailed classification frameworks, insights from managerial studies can more effectively be coupled with approaches from within psychology and anthropology.

@@

<!-- UNIVERSAL QUALITIES OF COMMUNITY LEADERS -->
There are some broad commonalities in the qualities of community leaders across culturally and economically diverse populations. When strong community leaders emerge, they are most typically high status males who embody the ideal qualities most respected by the group and necessary for successful leadership in the domains in which they have influence [@henrich_evolution_2001; @van_vugt_evolutionary_2006; @collier_politics_1981; @low_men_1992; @roscoe_social_2019]. There has been increasing interdisciplinary discussion and consensus on the closely related topics of social status, power, authority, rank, prestige, and leadership [e.g., @blader_whats_2014; @cheng_dominance_2019; @kantner_identifying_2010; @von_rueden_roots_2014]. We draw on discussions by @blader_whats_2014 and @von_rueden_roots_2014 and conceptualize social status as a relative indicator of an individuals social value by the group and their access to contested resources within the group. The relationship between leadership (i.e., social influence) and status (i.e., social value) is expected to be colinear [@von_rueden_leadership_2015; @van_vugt_genes_2020], however, if there is causality between the two it is likely to be complex, context-specific, and driven by underlying correlates.

<!-- MALE BIAS -->
Substantial evidence suggests there is a near universal male-bias in community, political, or group-level leadership across human societies [@low_sex_1992; @whyte_status_1978; @sanday_female_1981], as well as across diverse non-human social species [@smith_obstacles_2018; @tiger_female_1978]. @low_sex_1992 demonstrated in a systematic cross-cultural sample that political leaders of both sexes were present in only seven of 93 societies and in only two societies were female political leaders identified as equally powerful as their male counterparts. Even among these rare cases, powerful female political leaders are still infrequent within the group and women were more likely to hold influential leadership positions within kin groups [@low_sex_1992]. Empirical ethnographic studies have suggested sex-differences in political influence and leadership across human societies including among more egalitarian cultural contexts are more likely driven by ecological and social structural factors, rather than sex (or gender identity) per se, including social capital and access to same-sex allies [@yanca_female_2004] and/or sex differences in the sexual division of labor [@von_rueden_sex_2018]. At least two studies relying on ethnographic fieldwork comparing leaders and non-leaders in relativley small-scale societies revealed sex is not a strong predictor of community influence or elected leadership positions, after accounting for a broad range of demographic, phenotypic, cognitive, social, and other traits derived from the evolutionary theoretical literature [e.g., @garfield_investigating_2019; @von_rueden_sex_2018].

<!-- MORE UNIVERSAL QUALITIES FROM WEIRD AND NON WEIRD -->
Across cultural contexts elders, respected individuals, and highly intelligent or knowledgeable individuals tend to have greater influence over community decisions across cultures, including among the most egalitarian hunter-gatherers, stratified chiefdoms, and contemporary state-level organizations [@silverman_how_1978; @van_vugt_cognitive_2007; @tooby_cognitive_2006; @garfield_review_2019; @bass_handbook_1990]. Drawing on cross-national data through the GLOBE project from over 15,000 middle managers in businesses across 60 countries, @den_hartog_culture_1999 conclude that elements of charismatic and transformational leadership -- leadership styles when leaders express an inclusive vision of future progress, lead through inspiration, and develop initiatives in collaboration -- are universally endorsed across nations and are qualities perceived to be associated with effective leadership. In particular, inducing motivation, having foresight, being encouraging and communicative, trustworthiness, and building self-confidence among followers are suggested to be universal attributes of outstanding leadership in managerial contexts.

@@

<!-- NON WEIRD QUALITIES -->
Among small-scale societies with relative egalitarianism and local political autonomy community leaders, in addition to being older, respected males, tend to be individuals with large social networks and strong social support [@von_rueden_leadership_2014; @von_rueden_roots_2014; @walker_living_2012]. Physically, leaders across multiple contexts tend to be taller and more physically formidable. Such embodied capital might equip individuals for more effective leadership in the context of inter-group conflicts as well as in the context of intra-group dispute settlement and in management of collective actions [@lukaszewski_role_2016; @von_rueden_leadership_2014; @von_rueden_roots_2014]. Charisma and oratory skills are common properties of leaders, but are especially important in the context of egalitarianism where individual leaders must rely on persuasion and more subtle forms of influence to convince community members to come to agreement [@grabo_charismatic_2016].

>>>>

Across human societies, the influence of leaders stands to substantially shape the nature of group structure and group outcomes. Simultaneously, social norms, competing leaders, pressures from rival or cooperative groups, and followers preferences can greatly constrain a leader's agenda. The interactions between the strategies of individual leaders, pressures of followership, and group dynamics are likely to strongly contribute to organizational variation across human populations including social and economic trajectories [@wiessner_risk_1982] and cultural evolutionary processes [@henrich_big_2015].

Leadership and followership are critical in understanding group structure and have been the focus of much scientific attention across psychology, sociology, political science, and managerial studies. Until recently, this very large body of work was relatively disconnected from approaches to leadership in anthropology and biology. Under a unifying evolutionary framework, social and biological scientists have now produced a sizable body of theory on the evolution and cross-cultural patterning of leadership and followership [e.g., @price_evolution_2014; @henrich_big_2015; @richerson_tribal_2009; @van_vugt_evolutionary_2006; @pietraszewski_evolution_2019; @hooper_theory_2010; @kaplan_evolutionary_2009; @von_rueden_roots_2014; @hagen_leadership_2019], accompanied by a growing body of empirical results [e.g., @bowser_womens_2010; @cheng_two_2013; @garfield_investigating_2019; @garfield_evolutionary_2019; @von_rueden_leadership_2014; @von_rueden_sex_2018; @smith_leadership_2016; @smith_obstacles_2018]. This integration as been informative. Drawing on the study of leadership in evolutionary perspective and across diverse, small-scale and traditional societies can help develop more generalizable theories and offer applications to shape organizational structure in ways which are more consistent with human evolutionary history and evolved leader-follower psychologies [@von_rueden_leadership_2015; @garfield_review_2019; @van_vugt_selected_2010]. Nonetheless, much of this work has focused on community or political leaders, often overlooking the nature of leadership and followership across social domains and levels of social organization, such as among kin groups, economic groups, and sub-groupings of individuals within residential communities. Moreover, while the benefits leaders accrue for their services have received substantial attention the costs of leadership for both leaders and followers have been relatively under-investigated. Current outstanding questions for evolutionary leadership scholars include the degree to which the nature of leadership is associated with population-level variation at different levels of social organizations and the costs and benefits leaders and followers incur across distinct leadership systems.

Evolutionary scholars typically define leaders as individuals who maintain disproportionate influence over decision-making within groups [@von_rueden_leadership_2015], whatever the group may be. Leadership roles are also often associated with specific rights and responsibilities. Not surprisingly, the functions leaders serve then are often closely linked to group contexts and the salient challenges groups face. Certain types of groups, such as kin groups, economic groups, or political groups, may tend to favor leaders with specific qualities, however there can be substantial variation in the traits of leaders within these groups as well as between them.

We first briefly review theoretical, ethnographic, and empirical perspectives on the qualities of leaders, the functions they serve, and the costs and benefits of leadership across human societies, including WEIRD and non-WEIRD populations [i.e., @henrich_weirdest_2010]. We then outline the aims and methods of the current cross-cultural study, before presenting results and interpretations. Through novel exploratory analyses encompassing the full range of cultural diversity and diverse social contexts, we aim to advance leadership studies across disciplines and provide researchers with a foundation for more detailed approaches to human leadership across cultures and contexts.

@@

## The functions of leadership across cultures and contexts

Leaders emerge in all human societies. That is, within every population there are particular contexts in which one individual maintains a disproportionate level of influence in the group, functions to facilitate group goals, and provides services to the group. Adopting the broad definition of leadership given above, parents can also be defined as leaders of nuclear families. Strong community leadership, however, is extremely common cross-culturally including among hunter-gatherers and relatively small-scale societies [@garfield_review_2019; @von_rueden_leadership_2015].

Despite the universality of the phenomenon of leadership, the nature of leadership across populations and across social contexts is highly diverse and the emergence of leaders and the functions they serve are expected to be closely tied to group structure and group goals [@van_vugt_cognitive_2007]. Among many of the ethnographically described small-scale societies, which generally lack multi-level, institutionalized social and political structures, leadership at the community level is typically informal and ephemeral and leaders emerge contextually when demands of coordination or conflict resolution require a managerial presence [@von_rueden_leadership_2015]. In the absence of institutionalized organizational structure leaders often take on essential roles in group coordination and collective action, including facilitating conflict resolution within or between autonomous households or kin groups, managing individual investments to group-wide projects, and administering sanctions to individuals who do not appropriately contribute to collective actions, comply with group policies, or seriously violate social norms. @glowacki_leadership_2015 draw on empirical, ethnographic, and theoretical models of leader emergence and suggest in small-scale societies institutionalized leadership roles are likely to initially emerge to resolve conflicts between families. @roscoe_social_2009 draws on the ethnography of highland New Guinea in describing the modular structure of human societies and suggests the functions of community leaders in small-scale societies are structured by the joint demands of signaling military formidability to rival groups and facilitating within-group conflict resolution and cooperation. For "big-men" in these societies, in addition to military prowess, "talents for oratory, mediation, and organization were instrumental in reducing their internal conflict and enhancing the ability of their members to coordinate their actions as a group" [@roscoe_social_2009, p. 104].

Given demands of community organization and conflict resolution, leadership tends to be context-specific and leaders in such contexts often draw on persuasive capacities and leverage community-based respect, specialized skills or knowledge, and social support to achieve or maintain community influence [@kantner_identifying_2010; @lewis_leaders_1974; @boehm_egalitarian_1993]. The theoretical and ethnographic literature suggests these demands most commonly arise in the contexts of collective actions and in conflict resolution [e.g., @glowacki_leadership_2015; @gavrilets_convergence_2016; @hooper_theory_2010; @von_rueden_solving_2015; @henrich_big_2015; @pietraszewski_evolution_2019]. Demands of mediation in within-group conflict resolution, although variable in scale, are probably a universal feature of human groups [@king_origins_2009; @van_vugt_leadership_2008]. Demands of between-group interactions, including both conflict and cooperation are also very common across small-scale societies, including hunter-gatherers [@glowacki_evolutionary_2017; @hames_pacifying_2019; @richerson_cultural_2016; @henrich_evolution_2009; @apicella_social_2012]. Across relatively small-scale societies representing the group and managing both internal and external relationships and conflicts are expected to be the most salient functions of community leaders. Within communities, however, leaders may emerge whenever individuals congregate. The role of leaders in ritual contexts, such as in initiations [@singh_cultural_2017], and in providing social services, including facilitating marriages [@walker_evolutionary_2011] and organizing social activities [@wiessner_food_1996], are also expected to be common functions carried out by a leader.

Increased sedentarization and defensibility of resources is often associated with both social stratification and broader institutionalized managerial roles including stronger authority of community leaders and leadership beyond the community level [@johnson_evolution_1987; @johnson_organizational_1982; @kaplan_evolutionary_2009; @ames_hierarchies_1985; @panchanathan_indirect_2004]. Among horticulturalists and agriculturalists managerial leadership can promote more efficient use of shared resources such as water reserves and cooperative labor [e.g., @von_rueden_leadership_2014]. Pastoralists and coastal populations often face similar pressures concerning grazing lands and fishing access [e.g., @widmer_evolution_1988; @stevens_sherpa_1990]. In the wake of economic intensification, the functions of leadership becomes increasingly associated with managing economic systems and distribution of resources. Defensibility of critical resources creates increased opportunities for resource management and is often associated with increased territoriality and increased pressures of inter-group conflicts [@glowacki_evolutionary_2017]. Community leaders among populations more reliant on domesticated plant foods, livestock, and specific territories are often required to manage military forces and develop both offensive and defensive strategies in warfare, more so than mobile populations that rely on dispersed, less defensible resources [@lopez_evolutionary_2017].

In WEIRD societies, community leaders are primarily heads of governmental and corporate bodies [@bass_handbook_1990]. Other public service, nonprofit leaders and managers include law enforcement and police officers, college presidents and school principles, and grassroots community leaders, to name a few [@martiskainen_role_2017; @azzam_community_2003; @grace_school_2005]. These positions typically are associated with institutionalized responsibilities and specific functions. The structure of such organizations are generally hierarchical with nested bodies and an internal 'chain of command.' At higher levels of large-scale organizations, leaders primarily serve as representatives in inter-group interactions, facilitate the creation and adoption of policies, and serve as figure heads embodying the strategic vision of the organization [@keohane_thinking_2010]. They are not typically directly involved in resolving interpersonal conflicts between individuals, however, they are often the final end-point in a decision making hierarchy and the results of a single decision of these leaders can trickle down through industry or government with global ramifications.

Scholars in managerial studies have long discussed the behaviors of leaders and managers have developed multiple models on the functions of leadership. Classic management theorists [e.g., @davis_learning_1951] emphasized the planning, organizing, and controlling functions of leaders in formal organizational and business settings [see @bass_handbook_1990]. These perspectives focused on the rational, mechanistic aspects of group functioning and often overlooked interpersonal dynamics. Later theorists incorporated cognitive and socioemotional components of leadership functions. For example, @selznick_leadership_1957 describes the functions of organization leaders is to define the goals of the organization, create structures for accomplishing group goals, defend institutional integrity, and maintain internal consistency among group members.

@wofford_factor_1970 provides an early empirical view of managerial functions. A factor analysis of managerial behaviors from over 130 US employees' survey responses identified five interpretable, orthogonal factors: (1) an *order and group achievement* factor, identifying participative, democratic, and organizational planning function; (2) a *personal enhancement oriented* factor, associated with authority and control, (3) a *personal interaction* factor, associated with interpersonal relationships and employee support, (4) a *security and maintenance* factor, associated with self-preservation, and (5) a *dynamic and achievement oriented* factor, associated with aggressive, self-assured actions. Wofford's analysis provided a foundation for future empirical studies of leadership and management functions. He contrasted factor one type management, described as the "professional administrator" who effectively collaborates and integrates team members, from factors two and five, described as "autocratic leaders" strongly motivated by personal power and a willingness to rely on aggressive tactics [@wofford_factor_1970].

More recently, many managerial scholars have adopted integrative and comparative approaches to better understand cultural diversity in leadership systems. @behrendt_integrative_2017 draw on empirical results from the managerial literature and suggest the behaviors of leaders across managerial contexts are classifiable as either *task-oriented*, including enhancing understanding, strengthening motivation and facilitating implementation, or *relation-oriented*, including fostering coordination, promoting cooperation and mobilizing resources. This framework is intended to provide a more fine-grained and externally valid view of leadership functions, compared to traditional models of transactional leadership, which tend to oversimplify dimensions of leader-follower dynamics. More integrated and holistic models have also allowed managerial scholars to develop comparative theoretical frameworks for leader functioning. Culturally Endorsed Implicit Leadership Theory (CLT), suggests cultural variation in social norms has the potential to impact follower preferences for leadership styles [@moore_literature_2018]. The Global Leadership and Organizational Behavior Effectiveness (GLOBE) project is a large-scale international research program aimed at understanding leader effectiveness and cultural variation across large organizations. The GLOBE project has produced multiple cross-national data sets on leadership and hundreds of publications. For example, *The GLOBE Study of CEO Leadership Behavior and Effectiveness in 24 Countries* draws on data from over 1,000 CEOs and over 5,000 senior executives from corporations in varied industries across 24 countries [@house2013strategic]. The goals of the GLOBE CEO Study were to address (1) how national culture influences expectations for the behaviors of leaders, (2) which behaviors of CEOs are typically associated with organizational success, (3) differences between high-functioning and low-functioning CEOs, and (4) the importance of CEO behaviors aligning with societal expectations. Through path analysis the study concluded that cultural values indirectly predict CEO leadership behavior through culturally endorsed leadership expectations. That is, leaders moderate their behavior through culturally learned values and leadership-specific values.

The GLOBE CEO Study defines six "global leadership behaviors" developed from 21 "primary leadership behaviors." Among the global leadership behaviors, the GLOBE CEO Study found that charismatic leadership behavior is consistently the most important behavioral trait regarding firm performance and dedication of top management team (TMT) members, followed by team-oriented behavior and human-oriented behavior. Participative leadership was moderately related to TMT dedication, however was not related to firm performance. The most successful CEOs are those who are rated as more visionary, performance oriented, and decisive. Furthermore, results strongly suggest the fit between a CEOs behavioral profile and the expectations of leadership within a society strongly predict TMT dedication and firm performance. By higher-order features of leadership behavior and situating leadership functions within more detailed classification frameworks, insights from managerial studies can more effectively be coupled with approaches from within psychology and anthropology.

@@

There are some broad commonalities in the qualities of community leaders across culturally and economically diverse populations. When strong community leaders emerge, they are most typically high status males who embody the ideal qualities most respected by the group and necessary for successful leadership in the domains in which they have influence [@henrich_evolution_2001; @van_vugt_evolutionary_2006; @collier_politics_1981; @low_men_1992; @roscoe_social_2019]. There has been increasing interdisciplinary discussion and consensus on the closely related topics of social status, power, authority, rank, prestige, and leadership [e.g., @blader_whats_2014; @cheng_dominance_2019; @kantner_identifying_2010; @von_rueden_roots_2014]. We draw on discussions by @blader_whats_2014 and @von_rueden_roots_2014 and conceptualize social status as a relative indicator of an individuals social value by the group and their access to contested resources within the group. The relationship between leadership (i.e., social influence) and status (i.e., social value) is expected to be colinear [@von_rueden_leadership_2015; @van_vugt_genes_2020], however, if there is causality between the two it is likely to be complex, context-specific, and driven by underlying correlates.

Substantial evidence suggests there is a near universal male-bias in community, political, or group-level leadership across human societies [@low_sex_1992; @whyte_status_1978; @sanday_female_1981], as well as across diverse non-human social species [@smith_obstacles_2018; @tiger_female_1978]. @low_sex_1992 found in a systematic cross-cultural sample that in only seven of 93 societies political leaders of both sexes were present and in only two societies were female political leaders identified as equally powerful as their male counterparts. Even among these rare cases, powerful female political leaders are still infrequent within the group. Women were more likely to hold influential leadership positions within kin groups [@low_sex_1992]. More recently, empirical studies have suggested sex-differences in political influence and leadership across human societies including among more egalitarian cultural contexts are more likely driven by ecological and social structural factors, rather than sex (or gender identity) per se, including social capital and access to same-sex allies [@yanca_female_2004] and/or sex differences in the sexual division of labor [@von_rueden_sex_2018]. Ethnographic fieldwork comparing leaders and non-leaders in two distinct small-scale societies has revealed sex is not a strong predictor of community influence or elected leadership positions, after accounting for a broad range of demographic, phenotypic, cognitive, social, and other traits derived from the evolutionary theoretical literature [e.g., @garfield_investigating_2019; @von_rueden_sex_2018].

Across cultural contexts elders, respected individuals, and highly intelligent or knowledgeable individuals tend to have greater influence over community decisions across cultures, including among the most egalitarian hunter-gatherers, stratified chiefdoms, and contemporary state-level organizations [@silverman_how_1978; @van_vugt_cognitive_2007; @tooby_cognitive_2006; @garfield_review_2019; @bass_handbook_1990]. Drawing on cross-national data through the GLOBE project from over 15,000 middle managers across 60 countries, @den_hartog_culture_1999 conclude that elements of charismatic and transformational leadership -- leadership styles when leaders express an inclusive vision of future progress, lead through inspiration, and develop initiatives in collaboration -- are universally endorsed across nations and are qualities perceived to be associated with effective leadership. In particular, inducing motivation, having foresight, being encouraging and communicative, trustworthiness, and building self-confidence among followers are suggested to be universal attributes of outstanding leadership in managerial contexts.

@@

Among small-scale societies with relative egalitarianism and local political autonomy community leaders, in addition to being older, respected males, tend to be individuals with large social networks and strong social support [@von_rueden_leadership_2014; @von_rueden_roots_2014; @walker_living_2012]. Physically, leaders across multiple contexts tend to be taller and more physically formidable. Such embodied capital might equip individuals for more effective leadership in the context of inter-group conflicts as well as in the context of intra-group dispute settlement and in management of collective actions [@lukaszewski_role_2016; @von_rueden_leadership_2014; @von_rueden_roots_2014]. Charisma and oratory skills are common properties of leaders, but are especially important in the context of egalitarianism where individual leaders must rely on persuasion and more subtle forms of influence to convince community members to come to agreement [@grabo_charismatic_2016].

In the context of institutionalized social stratification, such as chiefdoms and many agricultural populations, community leadership is often highly influenced by heredity and formal class structures [@earle_how_1997; @stanish_evolution_2004; @redmond_introduction_1998]. Managerial elites, who maintain influence over community decisions are endowed with a high degree of social status rooted in ideologies and associated with their roles or kin group [@stanish_evolution_2010]. In such contexts community leaders often lack strong coercive authority, but maintain substantial influence through ideological systems and ritual or religious prowess inaccessible to non-elites [@stanish_evolution_2010]. The association of formal community leadership and supernatural abilities is likely an important precursor of institutionalized and coercive authority. Within state-level societies including non-industrial kingdoms leadership positions are proscribed and institutionalized. Ruling classes, parties, or political bodies monopolize social and political influence and maintain control through a variety of mechanisms including military force, ideological systems, and through the control of information and propaganda [@cohen_comparative_1967; @bodley_cultural_2011].
